---
title: "ivygapSE basics"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{ivygapSE basics}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Overview

The ivygapSE package includes molecular, imaging, and 
limited clinical data on glioblastoma (GBM) patients.

Expression data (RNA-seq) was developed in a complex design
involving tissue blocks and subblocks and a selection process.
See documents at https://vjcitn.github.io/ivygapSE for more
details.

To work with the available survival data in an elementary
way, the following steps can be taken.

# Basic setup

```{r genes}
library(ivygapSE)
library(MASS)
library(nlme)
library(DT)
data(ivySE)
# In this code we drop out the "CT"-only samples
struc = as.character(colData(ivySE)$structure_acronym)
spls = strsplit(struc, "-")
basis = vapply(spls, function(x) x[1], character(1))
specific = which(basis!= "CT")
iseSP = ivySE[ , specific]
useful = c("EIF4E3","ATP1A2","FGFR3","LMO3","NCAN","FXYD1","PSD2","PRODH",
           "HIF3A","HRSP12","KCNN3","PPM1K","KCNJ10","ADCYAP1R1","BMP7",
           "KAT2B","CNTN1","SAMD9L","SLC7A11","ECHDC2","FAM181B","SALL2","SASH1",
           
           "CCL4","CCL3","EGR3","IL1B","CCL2","GPR56","TNF","CX3CR1","SERPINE1","CSF1",
           "RAMP1","IL6ST","IL1A","PDE3B","CD276","FLT1","CXCL12","NR4A1","CSF1R","CCR1",
           "RHOB","SPHK1","SORL1","IL6R","VASH1","ADAM28","STAMBPL1","ADAMTSL2")
splim = iseSP[useful,]
cd = colData(splim)
stac = cd$structure_acronym
stac1 = sapply(strsplit(stac, "-"), "[[", 1)
table(stac1)
eexp = as.numeric(log(assay(splim["EIF4E3",])+1))
newdf = data.frame(eexp=eexp, id=splim$tumor_id, bl=splim$block_name)
# it would be nice to generalize so that LE and IT and other types of
# tissue classification could be modeled easily
```

## Enhance `newdf` so that it has binary representations of key sample types (CThbv, CTmvp, etc.)

We are now going to build up newdf so that it has binary outcomes
for different sample "types", i.e., leading edge, microvascular proliferation, etc.

```{r multioutcome}

newdf$isCTH = ifelse(stac1 == "CThbv", 1, 0)
table(newdf$isCTH)
      
newdf$isLE = ifelse(stac1=="LE", 1, 0)
table(newdf$isLE)

newdf$isCTMVP = ifelse(stac1=="CTmvp", 1, 0)  # microvascular proliferation
table(newdf$isCTMVP)

newdf$isCTPAN = ifelse(stac1=="CTPAN", 1, 0)  # pseudopalisading around necrosis
table(newdf$isCTPAN)
```

We can do a few more like above, pnz and IT.

## Define a mixed effect logistic regression tool

```{r definemodel}
fit_mixed_logistic = function( SE = splim, type = "isLE", gene, dataframe, verbose=FALSE ) {
  stopifnot(all(c("id", "bl", type) %in% names(dataframe)))
  stopifnot(gene %in% rownames(SE))
  dataframe[[gene]] = as.numeric(log(assay(SE[gene,])+1))
  fmla = as.formula(paste(type, "~", gene))
  glmmPQL(fmla, data=dataframe, fam=binomial, random=~1|id/bl, verbose=verbose)
  }
```

# Sample type prediction using mixed effect logistic regression

## Leading edge

```{r doLE}
LE_models = lapply(useful, function(x) fit_mixed_logistic(SE=splim, type="isLE", gene=x, dataframe=newdf) )
names(LE_models) = useful

coefs_LE= lapply(LE_models, function(x) summary(x)$tTable)


newtab_LE = data.frame(gene = names(coefs_LE),
                    effect = round(sapply(coefs_LE, function(x) x[2, "Value"]),3),
                    tstat = round(sapply(coefs_LE, function(x)x[2, "t-value"]), 3))


datatable(newtab_LE)
```

## Hyperplastic blood vessels

```{r doCTH}
CTH_models = lapply(useful, function(x) fit_mixed_logistic(SE=splim, type="isCTH", gene=x, dataframe=newdf) )
names(CTH_models) = useful

coefs_CTH= lapply(CTH_models, function(x) summary(x)$tTable)

newtab_CTH = data.frame(gene = names(coefs_CTH),
                    effect = round(sapply(coefs_CTH, function(x) x[2, "Value"]),3),
                    tstat = round(sapply(coefs_CTH, function(x)x[2, "t-value"]), 3))
datatable(newtab_CTH)
```

# Enhance the model to use pairs of genes

Now let's try pairs of genes.

```{r dopairs}
allpairs = combn(useful,2)
allpairs.list = lapply(1:ncol(allpairs), function(x) allpairs[,x])

fit_mixed_logistic_CT_two_genes = function( SE = splim, gene1, gene2, dataframe, verbose=FALSE ) {
  stopifnot(all(c("id", "bl", "isCTH") %in% names(dataframe)))  # so far just using CTH, but need to generalize FIXME
   stopifnot(gene1 %in% rownames(SE))
   stopifnot(gene2 %in% rownames(SE))
   dataframe[[gene1]] = as.numeric(log(assay(SE[gene1,])+1))
   dataframe[[gene2]] = as.numeric(log(assay(SE[gene2,])+1))
   fmla = as.formula(paste("isCTH", "~", gene1, "+", gene2))
   glmmPQL(fmla, data=dataframe, fam=binomial, random=~1|id/bl, verbose=verbose)
}

ff_CT_two_genes = lapply(allpairs.list[1:5], function(x) fit_mixed_logistic_CT_two_genes(SE=splim, 
                    gene1=x[1], gene2=x[2], dataframe=newdf) )
ff_CT_two_genes

coefs= lapply(ff_CT_two_genes, function(x) summary(x)$tTable)
coefs

gns_pair = lapply(coefs, function(x) rownames(x)[2:3])
gns_pair

gns = lapply(coefs, function(x) data.frame(genes=rownames(x)[2:3], effs=x[2:3, "Value"], tstats=x[2:3, "t-value"]))
gns
```
