---
title: "ivygapSE basics"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{ivygapSE basics}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---


The ivygapSE package includes molecular, imaging, and 
limited clinical data on glioblastoma (GBM) patients.

Expression data (RNA-seq) was developed in a complex design
involving tissue blocks and subblocks and a selection process.
See documents at https://vjcitn.github.io/ivygapSE for more
details.

To work with the available survival data in an elementary
way, the following steps can be taken.

```{r do1}
ii = rownames(installed.packages())
if (!("ivygapSE" %in% ii)) BiocManager::install("ivygapSE")
suppressPackageStartupMessages({
library(ivygapSE)
library(survival)
library(ggbiplot)
})
data(ivySE)
ivySE
```

This will show a SummarizedExperiment.  There are
270 RNA-seq samples, derived from 37 patients.
The variable `tumor_name` is used to identify different individual
donors to the project.

```{r lkta}
table(table(ivySE$tumor_name))
```

This shows that there are 2 tumors for which
18 RNA-seq studies were produced.  The RNA-seq
samples are likely from different subblocks and
may represent anatomically distinct components of
the tumor.  Our job is to help organize data
and analyses to help understand relationships
among many aspects of data collected in this 
study.

The first basic exercise is to assess the
prognostic information in a variable called
`mgmt_methylation`, collected on most tumors.
There is literature indicating that methylation
of the promoter of the MGMT gene is associated
with a more favorable survival profile.

Unfortunately the survival data is incomplete.
It can be obtained as follows.

```{r lktim}
metadata(ivySE)$tumorDetails$survival_days
```

We'll make a subset of the tumor data for those
with non-missing survival times.

```{r mkok}
td = metadata(ivySE)$tumorDetails
tdok = td[-which(is.na(td$survival_days)),]
```

Now we'll produce K-M plots.

Overall survival:
```{r lkov}
library(survival)
gbmsurv = Surv( tdok$survival_days, rep(1, nrow(tdok)) )
plot(survfit(gbmsurv~1))
```

Split by MGMT methylation
```{r met}
plot(survfit(gbmsurv~I(tdok$mgmt_methylation=="Yes")), lty=1:2)
```

## Biplot demo

```{r lkbip, out.width="250%"}
struc = as.character(colData(ivySE)$structure_acronym)
spls = strsplit(struc, "-")
basis = vapply(spls, function(x) x[1], character(1))

specific = which(basis!= "CT")

iseSP = ivySE[ , specific]

spbasis = basis[specific]

useful = c("EIF4E3", "RNASE4", "PBK", "RYR2", "TPD52L1",
    "GIMAP6", "PRKCB", "ATP6V1A", "RUNDC3B", "CCPG1")

splim = iseSP[useful,]

bb = prcomp(log(t(assay(splim))+1))
rownames(bb$x) = spbasis

biplot(bb, cex=c(.3,.6))
```

```{r lkn, out.width="150%"}
ggbiplot(bb, labels=rownames(bb$x), alpha=.4)
```
